### 2014-02-17 thesis-proposal/problem.tex
Updated proposal

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
Specifying and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that is built upon sound theoretical foundations, helps users identify and fix errors as early as possible, supports a natural programming style, and performs well across diverse problem domains and hardware platforms remains a grand challenge in computing. In view of this goal, researchers and domain experts (collectively, \emph{providers}) continue to develop new special-purpose syntax, static and dynamic semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these challenges in increasingly diverse contexts. Ideally, a provider would be able to develop and distribute these kinds of new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs. Unfortunately this is often infeasible because from the perspective of a library, the language's syntax and semantics are fully specified in advance, the compiler and run-time system are ``black box'' implementations of this fixed specification, and the other tools, like code editors and debuggers, operate according to fixed protocols. Providers of new system features must, as a result, take \emph{language-external approaches}, often by deriving a new programming system altogether. This approach has been encouraged, historically, by the availability of tools like compiler generators and language workbenches. We will argue that these approaches are technically problematic and that taking them has led to an unnecessary gap between research and practice. In their place, we will develop \emph{language-integrated extensibility mechanisms} that decentralize control over several core aspects of the programming system. By organizing new features around {types} and constraining them appropriately, we aim to show that these mechanisms can guarantee safety and non-interference of extensions while also being highly expressive.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

@@

When the semantics or implementation of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (supported by \emph{language workbenches} \cite{erdweg2013state},  \emph{DSL frameworks} \cite{fowler2010domain} or \emph{compiler generators} \cite{brooker1963compiler}), or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new languages. Many ``practical'' compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of these languages. Language-integrated mechanisms do not lead to such fragmentation.} or other tool. For example, a researcher interested in providing regular expression related features (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for these, perhaps basing it on an existing system containing some general-purpose features (\texttt{G}). A different researcher developing a new language-based parallel programming abstraction or implementation strategy (\texttt{P}) might  take the same strategy. A third researcher, developing an alternative parallel programming abstraction (\texttt{Q}) might again do the same. This results in a collection of distinct systems as diagrammed in Figure \ref{approaches}a. Unfortunately, when providers of new features take language-external approaches like this, it causes  problems for clients related to \textbf{orthogonality} and \textbf{client compatibility}, described below. %This latter method couples the semantics of the feature to the implementation details of a particular tool. Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.

@@

Recent evidence emphasizes the importance of this problem for driving research into practice: developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions and associated implementation strategies. This is problematic because different parallel programming abstractions or implementation strategies are more appropriate in different situations \cite{Tasharofi:2013rc}. Moreover,  parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting. Regular expression support, for example, may be simultaneously desirable for processing large amounts of textual data in parallel, but using these features together in the same compilation unit when they have been implemented by language-external means is difficult or impossible.  %Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.

@@

\paragraph{Client Compatibility} Even in cases where for each component of a system, there has been constructed a  system that is completely satisfactory, there remain problems at the interface between components. An interface that exposes the specialized constructs particular to one language (e.g. futures in a parallel programming language) cannot necessarily be safely and naturally consumed from another language (e.g. a general-purpose language). Tool support is also lost when calling into a different language. We call this fundamental issue the \emph{client compatibility problem}: code written by clients of a certain collection of features cannot always interface with code written by clients of a different collection  in a safe, performant and natural manner.

@@

This approach can work well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead. For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#. The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this highly-specialized invariant is left completely unchecked at language boundaries. 
%In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
In some cases, desirable features must be omitted entirely due to concerns about interoperability. F\#, for example, aimed to retain source compatibility with Ocaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no straightforward analogs in the type system of the CLI.
%\end{itemize}

\subsection{Language-Integrated Approaches}\label{language-integrated-approaches}
We argue that, due to these problems with orthogonality and client compatibility, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible. The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that have previously required central planning\footnote{One might compare today's programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies. Our safety constraints serve a role analagous to market regulation. We leave further development of this analogy to the reader.} so that language-external approaches are less frequently necessary. More specifically, we will show how control over aspects of the \textbf{syntax}, \textbf{static and dynamic semantics} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b. 
Such libraries have been called \emph{active libraries}  \cite{activelibraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features. Features implemented within active libraries can be imported as needed, unlike features implemented by external means, seemingly avoiding the problems of orthogonality and client compatibility.

>>>>

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
Specifying and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that is built upon sound theoretical foundations, helps users identify and fix errors as early as possible, supports a natural programming style, and performs well across diverse problem domains and hardware platforms remains a grand challenge in computing. In view of this goal, researchers and domain experts (collectively, \emph{providers}) continue to develop new special-purpose syntax, static and dynamic semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these challenges in increasingly diverse contexts. Ideally, a provider would be able to develop and distribute these kinds of new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs. Unfortunately this is often infeasible because from the perspective of a library, the language's syntax and semantics are fully specified in advance, the compiler and run-time system are ``black box'' implementations of this fixed specification, and the other tools, like code editors and debuggers, operate according to fixed protocols. Providers of new system features must, as a result, take \emph{language-external approaches}, often by deriving a new programming system altogether. This approach has been encouraged, historically, by the availability of tools like compiler generators and language workbenches. We will argue that these approaches are technically problematic and that taking them has led to an unnecessary gap between research and practice. In their place, we will develop \emph{language-integrated extensibility mechanisms} that decentralize control over several core aspects of the programming system. By organizing new features around {types} and constraining them appropriately, we aim to show that these mechanisms can guarantee safety and non-interference of extensions while remaining highly expressive.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

@@

When the semantics or implementation of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (supported by \emph{language workbenches} \cite{erdweg2013state},  \emph{DSL frameworks} \cite{fowler2010domain} or \emph{compiler generators} \cite{brooker1963compiler}), or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting semantics-preserving optimizations, should be considered a pernicious means for creating new languages. Many ``practical'' compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of these languages. Language-integrated mechanisms do not lead to such fragmentation.} or other tool. For example, a researcher interested in providing regular expression related features (let us refer to these collectively as \texttt{R}) might design a new system with built-in support for these, perhaps basing it on an existing system containing some general-purpose features (\texttt{G}). A different researcher developing a new language-based parallel programming abstraction or implementation strategy (\texttt{P}) might  take the same approach. A third researcher, developing an alternative parallel programming abstraction (\texttt{Q}) might again do the same. This results in a collection of distinct systems as diagrammed in Figure \ref{approaches}a. Unfortunately, when providers of new features take language-external approaches like this, it causes  problems for clients related to \textbf{orthogonality} and \textbf{client compatibility}. %This latter method couples the semantics of the feature to the implementation details of a particular tool. Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.

@@

Recent evidence indicates that this is one of the major barriers preventing research from being driven into practice: developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions and associated implementation strategies. This is problematic because different parallel programming abstractions or implementation strategies are seen as more appropriate in different situations \cite{Tasharofi:2013rc}. Moreover,  parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting. Regular expression support, for example, may be simultaneously desirable for processing large amounts of textual data in parallel, but using these features together in the same compilation unit when they have been implemented by language-external means is difficult or impossible.  %Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.

@@

\paragraph{Client Compatibility} Even in cases where for each component of a system there is a  system that is completely satisfactory, there remain problems at the interface between components. An interface that exposes the specialized constructs particular to one language (e.g. futures in a parallel programming language) cannot necessarily be safely and naturally consumed from another language (e.g. a general-purpose language). Tool support is also lost when calling into a different language. We call this fundamental issue the \emph{client compatibility problem}: code written by clients of a certain collection of features cannot always interface with code written by clients of a different collection  in a safe, performant and natural manner.

@@

This approach can work well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead. For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#. The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this highly-specialized invariant is left completely unchecked at language boundaries. 
%In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
In some cases, desirable features must be omitted entirely due to concerns about interoperability. F\#, for example, aimed to retain source compatibility with Ocaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no apparent analogs in the type system of the CLI.
%\end{itemize}

\subsection{Language-Integrated Approaches}\label{language-integrated-approaches}
We argue that, due to these problems with orthogonality and client compatibility, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible. The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give providers the ability to define, within libraries, new features that have previously required central planning
%\footnote{One might compare today's programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies. Our safety constraints serve a role analagous to market regulation. We leave further development of this analogy to the reader.}
so that language-external approaches are less frequently necessary. More specifically, we will show how control over aspects of the \textbf{syntax}, \textbf{static and dynamic semantics} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b. 
Such libraries have been called \emph{active libraries}  \cite{activelibraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features. Features implemented within active libraries can be imported as needed, unlike features implemented by external means, seemingly avoiding the problems of orthogonality and client compatibility.
